ODGI: scalable tools for pangenome graphs

### 2021-06-02 manuscript/main.tex
the Fig 1 doesn't shot a full length chromosome; pills of biology; little edits

    \abstract{ \textbf{Motivation:} Pangenomes graphs address some of the
      shortcomings of mainline genomics, in particular
      with regard to reference genome bias and identification of complex
      structural variants.
        %
        \\ \textbf{Results:} Here we present \odgi\, a novel range of
        scalable tools that implements new algorithms and has an efficient in-memory
        representation of DNA graphs. \odgi\ can handle hundreds of
        large genomes in memory with fast parallel execution. \odgi\ includes
        tools for detecting complex regions, extracting \textit{loci}, removing artifacts,
        exploratory analysis, and graph sorting, manipulation and visualisation.
        \odgi\ also provides an extensible API that allows for writing new tool.
        %
        \\
        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
        published under the permissive MIT license. Source code and documentation can be downloaded from
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        %
        \textbf{Contact:}
        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        egarris5@uthcs.edu
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME
        }
    \maketitle

@@

    Thanks to advances in sequencing technology, new full-length
    genome assemblies are produced at a high rate, including
    efforts from the Human Pangenome Reference Consortium (HPRC) and
    telomere-to-telemere projects, e.g.~\citep{Miga:2020}. In
    principle, a pangenome can model a full set of genomic elements of
    individuals in a given species or clade. Pangenomics contrasts
    with reference genome based approaches which relate sequences to a
    single consensus model of the genome~\citep{Eizenga:2020}. A
    pangenome can elegantly be represented by a graph data structure
    incorporating sequences as nodes and their connections as
    edges. These nodes are shared for homologs, paralogs and
    orthologs. So-called paths represent reads, contigs, haplotypes, individual chromosomes or part thereof, that run through the
    nodes.   A bi-directed graph can even contain both strands of DNA
    as paths. A graph-based approach is particularly relevant for
    exploring complex variation and highly repetitive regions.

    A major challenge is writing software that can deal with the sheer
    size of these graphs representing hundreds, and soon thousands, of
    human genomes. The \vg\ toolkit pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but does not scale for
    large data (see Table~\ref{tab:02}). We
    wrote a new toolset and API in \Cpp\ named Optimized Dynamic
    Genome/Graph Implementation (ODGI). \odgi\ implements a efficient
    graph structure in computer memory that can be dynamically updated using
    multiple CPU cores in parallel.

@@

    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
    provides convenient insight into genomic variation. \cmd{odgi viz}
    generates a linearized representation of the pangenome and is
    capable of handling full length human chromosomes
    (see Fig.~\ref{fig:1}a). \cmd{odgi sort} reorganises node order and
    simplifies the graph by applying several algorithms, including the
    novel `path-guided stochastic gradient descent' algorithm.
    \cmd{odgi layout} generates a layout file with the X-Y coordinates for
    all nodes and \cmd{odgi draw} draws it, bringing the pangenome representation
    in two dimensions.

    \topic{odgi stats, bin, depth \& degree} pangenome numerical
    counts provide insight into genome complexity in other
    ways. \cmd{odgi stats} returns the number of nodes,
    edges, paths, and graph length. \cmd{odgi bin} summarises path
    information into bins of a specified size, thereby enabling a
    summarised view of large genomes. \cmd{odgi depth} and
    \cmd{odgi degree} return the node depth and degree as defined by
    user provided criteria. These methods allow detection of complex
    genomic regions.

    \topic{odgi break, groom, chop \& unchop} pangenome graphs can be
    reorganised.  \cmd{odgi break} removes circular structures in the
    graph, thereby reducing complexity. \cmd{odgi groom} removes
    spurious inverted links by aligning the graph from the orientation
    that is supported by most paths. \cmd{odgi chop} divides long
    nodes into shorter ones at a maximum requested size, thereby
    simplifying downstream analysis. \cmd{odgi unchop} joins nodes
    and embedded sequences that do not change the graph topology,
    thereby obtaining a more compact representation of the graph.

@@

    \topic{odgi position} pangenome graphs are flexible when it comes
    to coordinate systems. \cmd{odgi position} can use the coordinate
    system from a contained reference genome --- a dynamic liftover
    --- to display coordinates and other localised features, as is shown in
    Fig.~\ref{fig:1b}. Note that multiple reference genomes can be contained
    in the graph and any contained path can be a reference.

@@

    Pangenome graphs stand to become a ubiquitous tool in
    genomics~\citep{Eizenga:2020}.  With \odgi\ we implemented a
    state-of-the-art tool suite, that can transform, analyse and
    visualise pangenome graphs at scale.  \odgi\ also exposes an extensible API
    that can be used by other tools, e.g., for genotyping, pangenome
    graph building and normalization. Future work will add support for
    RNA and protein sequences and expand on metadata capabilities of
    large pangenome graphs.

>>>>

    \abstract{ \textbf{Motivation:} Pangenomes graphs address some of the
      shortcomings of mainline genomics, in particular
      with regard to reference genome bias and identification of complex
      structural variants.
        %
        \\ \textbf{Results:} Here we present \odgi\, a novel range of
        scalable tools that implements new algorithms and has an efficient in-memory
        representation of DNA graphs. \odgi\ can handle hundreds of
        large genomes in memory with fast parallel execution. \odgi\ includes
        tools for detecting complex regions, extracting \textit{loci}, removing artifacts,
        exploratory analysis, and graph sorting, manipulation and visualisation.
        \odgi\ also provides an extensible API that allows for writing new tool.
        %
        \\
        \textbf{Availability:} \odgi\ is software written in the \Cpp\ programming language and
        published under the permissive MIT license. Source code and documentation can be downloaded from
        \href{https://github.com/pangenome/odgi}{https://github.com/pangenome/odgi}.\\
        %
        \textbf{Contact:}
        % \href{andreaguarracino@outlook.com}{andreaguarracino@outlook.com},
        egarris5@uthcs.edu
        \\
        \textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online. \FIXME
        }
    \maketitle

@@

    Thanks to advances in sequencing technology, new full-length
    genome assemblies are produced at a high rate, including
    efforts from the Human Pangenome Reference Consortium (HPRC) and
    telomere-to-telemere projects, e.g.~\citep{Miga:2020}. In
    principle, a pangenome can model a full set of genomic elements of
    individuals in a given species or clade. Pangenomics contrasts
    with reference genome based approaches which relate sequences to a
    single consensus model of the genome~\citep{Eizenga:2020}. A
    pangenome can elegantly be represented by a graph data structure
    incorporating sequences as nodes and their connections as
    edges. These nodes are shared for homologs, paralogs and
    orthologs. So-called paths can represent reads, contigs, haplotypes,
    individual chromosomes or part thereof, that run through the
    nodes. A bi-directed graph can even contain both strands of DNA
    as paths. A graph-based approach is particularly relevant for
    exploring complex variation and highly repetitive regions present
    which characterises the vertebrate genomes.

    A major challenge is writing software that can deal with the sheer
    size of these graphs representing hundreds, and soon thousands, of
    human genomes. The \vg\ toolkit pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but does not scale for
    large data (see Table~\ref{tab:02}). We
    wrote a new toolset and Application Programming Interface (API)
    in \Cpp\ named Optimized Dynamic Genome/Graph Implementation (ODGI).
    \odgi\ implements a efficient graph structure in computer memory
    that can be dynamically updated using multiple CPU cores in parallel.

@@

    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
    provides convenient insight into genomic variation. \cmd{odgi viz}
    generates a linearized representation of the pangenome (see Fig.~\ref{fig:1}a)
    and is capable of handling full length human chromosomes. \cmd{odgi sort}
    reorganises node order and simplifies the graph by applying several
    algorithms, including the novel 'path-guided stochastic gradient descent'
    algorithm. \cmd{odgi layout} generates a layout file with the X-Y
    coordinates for all nodes and \cmd{odgi draw} draws it, bringing
    the pangenome representation in two dimensions.

    \topic{odgi stats, bin, depth \& degree} pangenome numerical
    counts provide insight into genome complexity in other
    ways. \cmd{odgi stats} returns the number of nodes,
    edges, paths, and graph length. \cmd{odgi bin} summarises path
    information into bins of a specified size, thereby enabling a
    summarised view of large genomes. \cmd{odgi depth} and
    \cmd{odgi degree} return the node depth and degree as defined by
    user provided criteria. These methods allow the detection of complex
    genomic regions generated by highly repetitive sequences present in
    centromeres, segmental duplications, and acrocentric chromosomes.

    \topic{odgi break, groom, chop \& unchop} pangenome graphs can be
    reorganised. \cmd{odgi break} removes circular structures in the
    graph, thereby reducing its complexity. \cmd{odgi groom} removes
    spurious inverted links by aligning the graph from the orientation
    that is supported by most paths. \cmd{odgi chop} divides long
    nodes into shorter ones at a maximum requested size, thereby
    simplifying downstream analysis. \cmd{odgi unchop} joins nodes
    and embedded sequences that do not change the graph topology,
    thereby obtaining a more compact representation of the graph.

@@

    \topic{odgi position} pangenome graphs are flexible when it comes
    to coordinate systems. \cmd{odgi position} can use the coordinate
    system from a contained reference genome --- a dynamic liftover
    --- to display coordinates and other localised features, as is shown in
    Fig.~\ref{fig:1b}. Note that multiple reference genomes can be contained
    in the graph and any contained path can be used as a reference.

@@

    Pangenome graphs stand to become a ubiquitous tool in
    genomics~\citep{Eizenga:2020}. With \odgi\ we implemented a
    state-of-the-art tool suite that can transform, analyse and
    visualise pangenome graphs at large scale. \odgi\ also exposes an
    extensible API that can be used by other tools, e.g., for genotyping,
    pangenome graph building \citep{pggb} and normalization \citep {smoothxg}.
    Future work will add support for RNA and protein sequences and
    expand on metadata capabilities of large pangenome graphs.
